Improving Item-Item Similarity Estimation during Collaborative Filtering

### 2014-03-25 paper/paper.tex
Add first paragraph of Estimating... section.

In the first model, we imagine users are randomly presented with items to rate.
Given a sufficient period of time, all users will eventually rate all items such
that all elements of an item's vector correspond to a provided rating. The true
similarity in this model of {\em random} rating, which we denote $TrueSimR$, is
then the vector product of two item vectors. That is, 
\begin{align}
TrueSimR(A, B) = \frac{\sum\limits_{u\in U}
r_{u,A}r_{u,B}}{\sqrt{\sum\limits_{u\in U} r_{u,A}^2}
\sqrt{\sum\limits_{u\in U} r_{u,B}^2}}
\end{align}
where $U$ is the set of all users and $r_{u,I}$ is the rating given by user $u$
to item $I$.

In the second model, we imagine an unlimited supply of users each of whom
chooses to rate a subset of items. The lack of a rating is now meaningful since
it indicates a lack of interest of a particular user for a particular item. This
is captured by the model through a default rating of zero \footnote{We address
the issue of rating bias and the precise meaning of a zero valued rating later
in this report.}. At any point in time, we may estimate the true similarity in
this model of {\em preferential} rating, by computing the cosine similarity of
the two item vectors. Only in the limit of inifinite time does this estimate
converge to the true similarity, which we denote $TrueSimP$.  That is,
\begin{align}
TrueSimP(A, B) = lim_{t\to\infty}\frac{\sum\limits_{u\in U_{AB}}
r_{u,A}r_{u,B}}{\sqrt{\sum\limits_{u\in U_A} r_{u,A}^2}
\sqrt{\sum\limits_{u\in U_B} r_{u,B}^2}}
\end{align}
where $U_A$ is the set of users who have rated $A$, $U_B$ is the set of users
who have rated $B$, and $U_{AB}$ is the set of users who have rated both $A$ and
$B$.

>>>>

\section*{True Item-Item Similarity}

In the first model, we imagine users are randomly presented with items to rate.
Given a sufficient period of time, all users will eventually rate all items such
that all elements of an item's vector correspond to a provided rating. The true
similarity in this model of {\em random} rating, which we denote $TrueSimRand$,
is then the vector product of two item vectors. That is,
\begin{align}
TrueSimRand(A, B) = \frac{\sum\limits_{u\in U}
r_{u,A}r_{u,B}}{\sqrt{\sum\limits_{u\in U} r_{u,A}^2}
\sqrt{\sum\limits_{u\in U} r_{u,B}^2}},
\end{align}
where $U$ is the set of all users and $r_{u,I}$ is the rating given by user $u$
to item $I$.

In the second model, we imagine an unlimited supply of users each of whom
chooses to rate a subset of items. The lack of a rating is now meaningful since
it indicates a lack of interest of a particular user for a particular item. This
is captured by the model through a default rating of zero \footnote{We address
the issue of rating bias and the precise meaning of a zero valued ratings later
in this report.}. At any point in time, we may estimate the true similarity in
this model of {\em preferential} rating, by computing the cosine similarity of
the two item vectors. Only in the limit of inifinite time does this estimate
converge to the true similarity, which we denote $TrueSimPref$.  That is,
\begin{align}
TrueSimPref(A, B) = lim_{t\to\infty}\frac{\sum\limits_{u\in U_{AB}}
r_{u,A}r_{u,B}}{\sqrt{\sum\limits_{u\in U_A} r_{u,A}^2}
\sqrt{\sum\limits_{u\in U_B} r_{u,B}^2}},
\end{align}
where $U_A$ is the set of users who have rated $A$, $U_B$ is the set of users
who have rated $B$, and $U_{AB}$ is the set of users who have rated both $A$ and
$B$.

\section*{Estimating Item-Item Similarity}

Of course, most systems will never acquire ratings from all users for all items.
Nor are we able to work in the limit of infinite time. We must therefore make
estimates of the true item-item similarity using only partial information. In
the case of $TrueSimPref$, the cosine similarity computed at any point in time
serves as a simple and obvious estimate. In using this estimate, we are treating
all missing ratings as intentionally omitted. When estimating $TrueSimRand$, the
choice of how to handle missing ratings is perhaps less obvious. Following the
convention of Breese et al. \cite{Breese1998}, we choose to use the Pearson
correlation \footnote{The Pearson correlation is traditionally characterized by
subtracting an expectation value from each dimension. This is only approximately
true in our case. We discuss the how we subtract rating biases later in this
report.} for this estimate, ignoring all ratings by users who have not rated
both items. To summarize, our naive estimates are
\begin{align}
PearSim(A, B) = \frac{\sum\limits_{u\in U_{AB}}
r_{u,A}r_{u,B}}{\sqrt{\sum\limits_{u\in U_{AB}} r_{u,A}^2}
\sqrt{\sum\limits_{u\in U_{AB}} r_{u,B}^2}}
\stackrel{?}{\approx} TrueSimRand(A, B)
\end{align}

\begin{align}
CosSim(A, B) = \frac{\sum\limits_{u\in U_{AB}}
r_{u,A}r_{u,B}}{\sqrt{\sum\limits_{u\in U_A} r_{u,A}^2}
\sqrt{\sum\limits_{u\in U_B} r_{u,B}^2}}
\stackrel{?}{\approx} TrueSimPref(A, B).
\end{align}
